Bayesian Deep Gaussian Processes for Correlated Functional Data: \\ A Case Study in Cosmological Power Spectra

### 2025-06-26 paper/main.tex
edits to 4.1

Let $\psi \in \mathbb{R}^{p_\psi}$ denote the set of parameters that define a particular cosmology.
Mira-Titan has 8 such parameters as described in Section \ref{sec:data}; CAMB has 6.
Our goal is to effectively predict the power matter spectrum as a function of $\psi$, i.e. $\hat{S}(\psi)$, 
for a held-out cosmology.  We observe $m$-many training cosmologies and fit our DGP.FCO 
model (Section \ref{sec:inference}) to each one.  Let $\hat{S}_j$ for $j \in \{1,\ldots,m\}$ denote
the estimated posterior mean of each training cosmology.  In total, our training data comprises
the $p_\psi$-dimensional input $\psi_j$ and the $n$-dimensional output $\hat{S}_j$ (which is a 
function of $x_i = \log_{10}(k_i)$) for $j \in \{1,\ldots,m\}$.

To handle the functional nature of this output, we reformulate the problem using principal
components \citep[PCs, e.g.][]{banerjee2014linear}.  
We first center all $\hat{S}_j$ by subtracting $\bar{S} = \frac{1}{m}\sum_{j=1}^{m} \hat{S}_j$, 
then we combine them into an $n\times m$ matrix $\boldsymbol\eta$.  We use the first $p_\eta$ 
principal components to decompose $\boldsymbol\eta = B\Gamma^\top$ where $B$ is the 
$n\times p_\eta$ matrix of basis functions and $\Gamma$ is the $m\times p_\eta$ matrix of basis weights.  
Details on this decomposition are reserved for \ref{sec:apdx_basis}. For both the CAMB and Mira-Titan 
datasets, we set $p_\eta=10$. As a visual, the left panel of Figure \ref{fig:mean_PCs_oneW} shows 
$\bar{S}$ for the Mira-Titan training cosmologies.  The center panel shows the $p_\eta$ basis functions.

@@

Let $\gamma_i^{(j)}$ denote the $ij^\textrm{th}$ element of $\Gamma$, representing the basis weight for the 
$j^\textrm{th}$ cosmology and the $i^\textrm{th}$ PC.
Conditional on $B$, our training data now consists of scalars $\gamma_i^{(j)}$ as a function of $\psi_j$
for $j \in \{1,\ldots,m\}$ and $i \in \{1,\ldots,p_\eta\}$.  If we can effectively predict these basis weights
as a function of $\psi$ (i.e., $\hat{\gamma}_i(\psi)$ for $i \in \{1,\ldots,p_\eta\}$), then we may obtain predictions 
of the power spectrum as
\[
\hat{S}(\psi) = \sum_{i=1}^{p_\eta} \hat{\gamma}_i(\psi) B_i,
\]
where $B_i$ is the $i^\textrm{th}$ basis function (column) of $B$.

@@

In alignment with \cite{higdon2010estcosmo} and \cite{moran2023mira}, we use 
GPs to model the principal components' weights. 
We assume a powered exponential kernel for the Gaussian process and employ 
\texttt{GPfit} to perform a multi-start gradient based search for GP's hyperparameters 
\citep{macdonald2015gpfit}. That is, the $i$th vector of PC weights $\gamma_i(\psi)$, 
$i \in \{1,\ldots,p_\eta\}$, will be modeled as a zero mean GP: 
$\gamma_i(\psi) \sim \mathrm{GP}(0, \sigma^2R)$, with

\begin{equation*}
    R_{uv} = \prod_{j=1}^{p_\psi}\exp\left\{-10^{\beta_j}|\psi_{uj}-\psi_{vj}|^\alpha\right\}.
\end{equation*}

We fix $\alpha=1.95$ and estimate $\sigma^2$ and each $\beta_j$ with $j \in \{1,...,p_\psi\}$, and repeat this
process for all of the $p_\eta$ PCs. Using these trained GPs from the PCs, we predict the spectra for 
each of the $N$ held-out cosmologies (such as $\psi^*$) for both the CAMB and Mira-Titan datasets.

We fix the power parameter $\alpha = 1.95$ and estimate the lengthscale parameters $\beta_1, \ldots, \beta_{p_\psi}$ 
from the data using maximum likelihood. To obtain the predicted basis weight $\hat{\gamma}_i(\psi^*)$ 
for a new cosmology $\psi^* \in \mathbb{R}^{p_\psi}$, 
we use the posterior mean from a zero-mean GP model fit to the training data 
$\{(\psi_j, \gamma_i^{(j)})\}_{j=1}^m$. Under a zero-mean GP, the posterior mean is given by:

@@

where $\boldsymbol{\gamma}_i = [\gamma_i^{(1)}, \gamma_i^{(2)}, \ldots, \gamma_i^{(m)}]^\top$ is the vector of 
$m$ observed PC weights for the $i$th component, $R$ is the $m \times m$ correlation matrix between 
all pairs of training inputs defined above, and $r(\psi^*) \in \mathbb{R}^m$ is the vector of correlations between 
the test input $\psi^*$ and each training input $\psi_j$. To illustrate, in the right panel of Figure \ref{fig:mean_PCs_oneW}, 
we show how the predicted GP mean for the weight of the fifth PC ($\gamma_5$) for Mira-Titan 
varies as a function of the first cosmological parameter $\psi_1$, while the other parameters are fixed at a (scaled)
midpoint value of 0.5. With these estimates $\hat{\gamma}(\psi^*)$, we predict power spectra at 
held-out cosmologies for both the CAMB and Mira-Titan datasets.

>>>>

Let $\psi \in \mathbb{R}^{p_\psi}$ denote the set of parameters that define a particular cosmology.
Mira-Titan has 8 such parameters as described in Section \ref{sec:data}; CAMB has 6.
Our goal is to effectively predict the power matter spectrum as a function of $\psi$, i.e., $\hat{S}(\psi)$, 
for a held-out cosmology.  We observe $m$-many training cosmologies and fit our DGP.FCO 
model (Section \ref{sec:inference}) to each one.  Let $\hat{S}_j$ for $j \in \{1,\ldots,m\}$ denote
the estimated posterior mean of each training cosmology.  In total, our training data comprises
the $p_\psi$-dimensional input $\psi_j$ and the $n$-dimensional output $\hat{S}_j$ (which is a 
function of $x_i = \log_{10}(k_i)$) for $j \in \{1,\ldots,m\}$.

To handle the functional nature of this output, we reformulate the problem using principal
components \citep[PCs; e.g.,][]{banerjee2014linear}.  
We first center all $\hat{S}_j$ by subtracting $\bar{S} = \frac{1}{m}\sum_{j=1}^{m} \hat{S}_j$, 
then we combine them into an $n\times m$ matrix $\boldsymbol\eta$.  We use the first $p_\eta$ 
principal components to decompose $\boldsymbol\eta = B\Gamma^\top$ where $B$ is the 
$n\times p_\eta$ matrix of basis functions and $\Gamma$ is the $m\times p_\eta$ matrix of basis weights.  
Details on this decomposition are reserved for \ref{sec:apdx_basis}. For both the CAMB and Mira-Titan 
datasets, we set $p_\eta=10$. As a visual, the left panel of Figure \ref{fig:mean_PCs_oneW} shows 
$\bar{S}$ for the Mira-Titan training cosmologies.  The center panel shows the $p_\eta$ basis functions.

@@

Let $\gamma_i^{(j)}$ denote the $ij^\textrm{th}$ element of $\Gamma$, representing the basis weight for the 
$j^\textrm{th}$ cosmology and the $i^\textrm{th}$ PC.
Conditional on $B$, our training data now consists of scalars $\gamma_i^{(j)}$ as a function of $\psi_j$
for $j \in \{1,\ldots,m\}$ and $i \in \{1,\ldots,p_\eta\}$.  If we can effectively predict these basis weights
as a function of $\psi$ (i.e., $\hat{\gamma}_i(\psi)$ for $i \in \{1,\ldots,p_\eta\}$), then we may obtain predictions 
of the power spectrum as
\begin{equation}\label{eq:spred}
\hat{S}(\psi) = \sum_{i=1}^{p_\eta} \hat{\gamma}_i(\psi) B_i,
\end{equation}
where $B_i$ is the $i^\textrm{th}$ basis function (column) of $B$.

@@

In alignment with \cite{higdon2010estcosmo} and \cite{moran2023mira}, we use 
GPs with a power exponential kernel to model the principal components' weights. 
That is, each vector of PC weights $\gamma_i(\psi)$, $i \in \{1,\ldots,p_\eta\}$, 
will be modeled as a zero mean GP: $\gamma_i(\psi) \sim \mathrm{GP}(0, \sigma^2R)$, with kernel
\begin{equation*}
    R^{u,v} = \prod_{j=1}^{p_\psi}\exp\left(-10^{\beta_j}|\psi_{uj}-\psi_{vj}|^\alpha\right).
\end{equation*}
We fix $\alpha=1.95$ and estimate $\sigma^2$ and each lengthscale $\beta_1,\ldots,\beta_{p_\psi}$ 
with the \texttt{GPfit} {\sf R}-package \citep{macdonald2015gpfit}, which employs multi-start 
gradient-based maximum likelihood estimation.  We repeat this estimation
process for all of the $p_\eta$ PCs.  Conditioned on kernel hyperparameters,
posterior predictions of $\hat{\gamma}_i(\psi^*)$ for a new cosmological parameterization 
$\psi^* \in \mathbb{R}^{p_\psi}$ follows:

@@

where $\boldsymbol{\gamma}_i = [\gamma_i^{(1)}, \gamma_i^{(2)}, \ldots, \gamma_i^{(m)}]^\top$ 
is the vector of $m$ observed PC weights for the $i^\textrm{th}$ component, 
$R$ is the $m \times m$ correlation matrix between 
all pairs of training inputs defined above, and $r(\psi^*) \in \mathbb{R}^m$ is 
the vector of correlations between the test input $\psi^*$ and each training input 
$\psi_j$. To illustrate, the right panel of Figure \ref{fig:mean_PCs_oneW} shows
the predicted weight of the fifth PC ($\gamma_5$) for Mira-Titan as a function 
of the first cosmological parameter ($\psi_1$), while the other parameters are fixed 
at their midpoint.  With all estimated PC weights, $\hat{\gamma_i}(\psi^*), i=1,\ldots,p_\eta$, 
we predict power spectra following Eq.~\ref{eq:spred}.
